Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-01-22 script/manuscript_MEE.Rmd
adding in more cleanly processed post class seqs

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Reed Clark Benkendorf$^1$$^,$$^2$, Emily J. Woodworth$^1$$^,$$^2$, Jane E. Ogilive$^3$$^,$$^4$,   
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods.  
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/blast_seqs_classified.csv') %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_Seqs_Class = sum(Prcnt_seqs)) %>% 
  ungroup()
blst_seqs <- distinct(blst, sample.id)

@@

blst_mertensia <- blst %>% 
  filter(Genus == 'Mertensia') %>% 
  select(sample.id,  Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_seqs, Prcnt_Seqs_Class) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_seqs = sum(Prcnt_seqs),
         morphotype = 'MERTENSIA') %>% 
  distinct(sample.id, .keep_all = T)

@@

```{r clean sequence numbers from environment}
rm( krak, brack, blast, loci, no_loci, aa, bb, cc, dd)
```

@@

To detect whether the sequencing reads were semi-quantitative the proportion of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia* Roth, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then *corrected* to constitute the entire sample.

@@

Sequences which were assigned to genera without any species predicted by spatial analyses, were investigated by hand after post-processing steps. These were manually assigned to a variety of ranks, occasionally to genus #COUNTS# and family #COUNTS#,  by consultation of the alpha-taxonomic literature (@sadeghian2015molecular, @sennikov2017phylogenetic, @rabeler2016new, @pusalkar2015taxonomic, @moore2003its,  @weber1998new).

@@

blst <- read.csv('../data/blast_seqs_classified.csv') %>% 
  distinct(sample.id)

@@

tb <- blst %>% 
  left_join(., blst_cnt, by = 'species') %>% 
  group_by(species, name) %>% 
  mutate(Mu_Seqs = sum(Prcnt_seqs)/n,
         species = str_remove(species, 'Bombus '))  %>% 
  distinct(species, name, .keep_all = T) %>% 
  ungroup(name) %>% 
  slice_max(Mu_Seqs, n = 10) %>% 
  select(plant.species = name, species, n, Mu_Seqs, Genus, Family) %>% 
  mutate(rank = 1:n()) %>% 
  ungroup() %>% 
  mutate(Family = case_when(
    Genus %in% c('Trollius', 'Caltha', 'Aquilegia', 'Delphinium') ~ 'Ranunculaceae',
    Genus == 'Hydrophyllum' ~ 'Hydrophyllaceae',
    Genus == 'Epilobium' ~ 'Onagraceae',
    Genus == 'Viola' ~ 'Violaceae',
    Genus == 'Salix' ~ 'Saliceae',
    Genus == 'Cynoglossum' ~ 'Boraginaceae', 
    Genus == 'Arenaria' ~ 'Caryophyllaceae',
    Genus == 'Parnasiia' ~ 'Celastraceae',
    TRUE ~ Family 
  ))

@@

>>>>

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Reed Clark Benkendorf$^1$$^,$$^2$, Emily J. Woodworth$^1$$^,$$^2$, Jane E. Ogilive$^3$$^,$$^4$,   
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods.  
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/Post_Classified_BLAST_sqs.csv') %>% 
  rename(sample.id = Sample, Family = FAMILY, Genus = GENUS) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_Seqs_Class = sum(Prcnt_seqs)) %>% 
  ungroup()
blst_seqs <- distinct(blst, sample.id)

@@

blst_mertensia <- blst %>% 
  filter(Family == 'Boraginaceae') %>% 
  select(sample.id,  Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_Seqs_Class) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_seqs = sum(Prcnt_seqs),
         morphotype = 'MERTENSIA') %>% 
  distinct(sample.id, .keep_all = T)

@@

```{r Read Post Processing}

pp_seqs_t <- read.csv( file.path('../data', 'Sequence_Reclass_table.csv') )

knitr::kable(pp_seqs_t, align = "c", 
             col.names = c('Condition', 'No. Class.', 'Prcnt. Class.', 'Total Seqs', 'Rank'),
             caption = "Post classification of Sequences via Taxonomy and Ecology") %>% 
  kableExtra::kable_styling(
      latex_options = "striped",
      stripe_color = "gray!10")

seqs_sp <- sum (pp_seqs_t[pp_seqs_t$Rank=='Species','Prcnt_Records'] )
seqs_ge <- sum (pp_seqs_t[pp_seqs_t$Rank=='Genus','Prcnt_Records'] )
seqs_fa <- sum (pp_seqs_t[pp_seqs_t$Rank=='Family','Prcnt_Records'] )

sp_total_seqs <- sum (pp_seqs_t[pp_seqs_t$Rank=='Species','Total_Prcnt_Seqs'] )
sp_total_seqs <- sum (pp_seqs_t[pp_seqs_t$Rank=='Genus','Total_Prcnt_Seqs'] )
fa_total_seqs <- sum (pp_seqs_t[pp_seqs_t$Rank=='Family','Total_Prcnt_Seqs'] )
```

The initial classification of sequences which were made by BLAST were reviewed programmatically, using predicted presence of the species (from spatial modelling), modelled flowering time (from temporal modelling), and taxonomy (from existing sources). A sequential process was utilized which reassigned sequences based on binary combinations of the factors above (Appendix XX). Given the relative sparsity of the number, and relatedness, of species represented in the sequence database this was performed to: 1) Identify locally present species represented by surrogates in the DB 2) Reduce false classifications of focal species 3) Identify high confidence sequence matches. Each of the top ten taxa which were identified by BLAST from the aforementioned process composed `r sum(pp_seqs_t$Samples_Classified)` distinct records (CONFUSING WORDS), of these `r seqs_sp`% of the reads were classified to species `r sp_total_seqs`% which represent of the total classified sequences, `r seqs_ge`% of the reads were classified to genus representing `r sp_total_seqs`% of the classified reads, and `r seqs_fa`% of the records were classified to genus `r fa_total_seqs`% of total sequences.

*Sequences which were assigned to genera without any species predicted by spatial analyses, were investigated by hand after post-processing steps. These were manually assigned to a variety of ranks, occasionally to genus #COUNTS# and family #COUNTS#, by consultation of the alpha-taxonomic literature (@sadeghian2015molecular, @sennikov2017phylogenetic, @rabeler2016new, @pusalkar2015taxonomic, @moore2003its,  @weber1998new). * <- went back on this as new function captures most of it, but don't throw out.

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`. Both morphotypes of Asteraceae pollen `r toString(dd)`, and Ericaceae `r toString(ee)`.

```{r clean sequence numbers from environment}
rm( krak, brack, blast, loci, no_loci, aa, bb, cc, dd, fa_total_seqs, 
    ge_total_seqs, sp_total_seqs, seqs_sp, seqs_ge, seqs_fa)
```

@@

To detect whether the sequencing reads were semi-quantitative the subset of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia*, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then relativized to constitute the entire sample.

@@

@@

blst <- read.csv('../data/Post_Classified_BLAST_sqs.csv') %>% 
  rename(sample.id = Sample, Genus = GENUS, Family = FAMILY) %>% 
  distinct(sample.id)

@@

tb <- blst %>% 
  left_join(., blst_cnt, by = 'species') %>% 
  group_by(species, name) %>% 
  mutate(Mu_Seqs = sum(Prcnt_seqs)/n,
         species = str_remove(species, 'Bombus '))  %>% 
  distinct(species, name, .keep_all = T) %>% 
  ungroup(name) %>% 
  slice_max(Mu_Seqs, n = 10) %>% 
  select(plant.species = name, species, n, Mu_Seqs, Genus, Family) %>% 
  mutate(rank = 1:n()) %>% 
  ungroup() %>% 
  mutate(Family = case_when(
    Genus %in% c('Trollius', 'Caltha', 'Aquilegia', 'Delphinium') ~ 'Ranunculaceae',
    Genus == 'Hydrophyllum' ~ 'Hydrophyllaceae',
    Genus == 'Epilobium' ~ 'Onagraceae',
    Genus == 'Viola' ~ 'Violaceae',
    Genus == 'Salix' ~ 'Saliceae',
    Genus == 'Cynoglossum' ~ 'Boraginaceae', 
    Genus == 'Arenaria' ~ 'Caryophyllaceae',
    Genus == 'Parnassia' ~ 'Celastraceae',
    Genus %in% c('Erigeron', 'Symphyotrichum') ~ 'Asteraceae',
    TRUE ~ Family 
  ))

@@

APPENDIX XX - Review Classified Reads and Reassign

```{r}
knitr::include_graphics('../graphics/tables/reclassify_reads_table.pdf')
```
